Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-05-16 paper.tex
Update on Overleaf.

\documentclass[11pt]{article}
\input{header.tex}

\date{}
\renewcommand\abstractname{\fontsize{14pt}{0}\textbf{Abstract}\selectfont}

\usepackage[left=25mm, right=25mm, top=25mm, bottom=25mm, includehead=false, includefoot=false]{geometry}

\usepackage{graphicx}
\usepackage{url}
\usepackage[round,semicolon]{natbib}  % Citation styles https://www.sharelatex.com/learn/Natbib_citation_styles
\bibliographystyle{humannat}
\renewcommand{\bibsection}{}
\renewcommand{\bibhang}{\setlength{-1px}}

\usepackage{authblk} % For author lists
\renewcommand\Authfont{\fontsize{11}{1}\selectfont}
\renewcommand\Affilfont{\fontsize{9}{1}\selectfont}

\renewcommand*\footnoterule{}

\usepackage[table]{xcolor}
\usepackage[parfill]{parskip} % Line between paragraphs
\usepackage{amsmath}
\pagenumbering{arabic}

\usepackage{sectsty}
\allsectionsfont{\sffamily}

\usepackage[pdftex]{hyperref} 
\hypersetup{pdfborder={0 0 0} }

@@

\title{\sffamily\fontsize{16}{0}\textbf{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}}

\author[1]{Cl{\' e}mentine Cottineau\thanks{}}
\author[2,4]{Juste Raimbault}
\author[3]{Marion Le Texier}
\author[4]{Florent Le N{\' e}chet}
\author[2,5]{Romain Reuillon}
\affil[1]{Centre for Advanced Spatial Analysis, University College London, UK}
\affil[2]{UMR 8504 G{\' e}ographie-cit{\' e}s, Paris, France}
\affil[3]{UMR 6266 IDEES, Universit{\' e} de Rouen Normandie, France}
\affil[4]{Laboratoire Ville Mobilit{\' e} Transport, Universit{\' e} Paris-Est, France}
\affil[5]{Instit, ut des Syst{\` e}mes Complexes Paris Ile-de-France, France}

\affil[*]{\texttt{, Email: c.cottineau@ucl.ac.uk }}

@@

\begin{abstract}
\noindent
\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust. Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment. In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models. We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves. We do so using high performance computing in a very simple and straightforward open-source workflow.
$ $ \\ {\bf Keywords:} Space, Initial conditions, Sensitivity, ABM.

\end{abstract},

@@

\section{Introduction}

@@

\subsection{Why spatial patterns are expected to impact social simulations}

@@

\subsection{Objective}

@@

% **************

\section{Previous attempts in the literature}

\subsection{In general}

Paragraph not as detailed as the following one, but just to illustrate the fact that the issue is an important concern in different types of spatial simulation modelling approaches?

Eg. : effects of delineation and scale of cities in LUTI models \citep{Thomasetal2017}, etc.

\subsection{For Schelling model in particular}

@@

\subsection{And for SugarScape?}

% **************

\section{Method}

@@

\begin{figure}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure} %

\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure? By graphically showing on which parts of the figure the different subsections focus? Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}

\subsection{Constructing a spatial generator}

@@

\begin{figure}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure} %

\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}

@@

\subsection{Comparing Phase Diagrams}

@@

where $f_{\vec{alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.

% ************** 
\section{Application cases}

@@

\subsection{Schelling's model of residential segregation}

@@

\subsection{Sugarscape model of resource extraction and population settlement}

@@

% ************** 
\section{Results}

@@

We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results. We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.

% ************** 
\section{Discussion}

% **************

@@

\section{Acknowledgements}

The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1. Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.

\section{References}

\bibliography{spacematters}

>>>>

\documentclass[Afour,sageh,times]{sagej}
\input{header.tex}

@@

\runninghead{Cottineau \& al.}

\title{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}

\author{Cl{\' e}mentine Cottineau\affilnum{1}, Juste Raimbault\affilnum{2,4}, Marion Le Texier\affilnum{3}, Florent Le N{\' e}chet\affilnum{4}, Romain Reuillon\affilnum{2,5}
}

\affiliation{\affilnum{1}Centre for Advanced Spatial Analysis, University College London, UK\\
\affilnum{2}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
\affilnum{3}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
\affilnum{4}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
\affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}

\corrauth{Corresponding author
address.}

\email{c.cottineau@ucl.ac.uk}

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust. Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment. In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models. We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves. We do so using high performance computing in a very simple and straightforward open-source workflow.
\end{abstract}

\keywords{Space, Initial conditions, Sensitivity, ABM}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why spatial patterns are expected to impact social simulations}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Objective}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Previous attempts in the literature}
%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
\subsection{In general}

The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}. Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same. On the contrary, the scarcity of these considerations in the geographic simulation model literature questions the generalisation of their results, as it has for instance been showed in the case of LUTI models \citep{Thomasetal2017}, of diffusion processes using ABM \citep{LeTexierCaruso2017}.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{For Schelling model in particular}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{And for SugarScape?}

%%%%%%%%%%%%%%%%%%%%%%
\section{Method}
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{General workflow of our method\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure? By graphically showing on which parts of the figure the different subsections focus? Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}} \label{fig:method} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructing a spatial generator}

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing Phase Diagrams}

@@

where $f_{\vec{\alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.

\todo[JR]{explain this in clear words}

%%%%%%%%%%%%%%%%%%%%%%
\section{Application cases}
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Schelling's model of residential segregation}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sugarscape model of resource extraction and population settlement}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Schelling's model}

We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results. We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sugarscape model}

For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.

%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
% Acknowledgements

\begin{acks}
The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1. Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
\end{acks}

%%%%%%%%%%%%
%% References
%%%%%%%%%%%%

\bibliographystyle{SageH}
\bibliography{spacematters,biblio}
